Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-15 ms.tex
revised model examples and intro

According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e. mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated. 
The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
These separate principles were recently unified by~\citet{Altenberg2017}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism in which the mutation rate increases in individuals with poor fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population size and constant-viability selection in which fitness is determined by the number of mutant alleles accumulated in the genotype.
The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (I-C+MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

The types $A_k$ can represent a single or multiple haploid genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and any horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
Importantly, different types can have different rates; that is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

In the following section we present three examples of the model presented in Eq.~\ref{eq:model} for mutation and migration.

@@

Here we consider a large population of haploids and a trait determined by a major locus with $n$ possible alleles $A_1, A_2, ..., A_n$.
The mutation rates $\mu_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-\mu_k$, the allele $A_k$ does not mutate, and with probability $\frac{\mu_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $\mu_k$, to depend on the identity of the allele $A_k$.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_k{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = (1 - \mu_k) \frac{w_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{\mu_j \frac{w_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{w_k f_k}$ is the population mean fitness.

This model can be rephrased in terms of the general model: 
Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
\begin{equation} 
C_k = \mu_k, \quad
D_k = w_k, \quad
M_{i,j} = \begin{cases}
\frac{1}{n-1}, & i \ne j \\
0, & i=j
\end{cases}.
\end{equation}
Note that here $M$ is irreducible and therefore so is $(I-C+MC)D$.

@@

Here we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Importantly, the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= (1 - \beta m_n) \frac{w_n}{\bar{w}} f_n + \delta m_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k=0}^{n}{w_k f_k}$ is the population mean fitness.

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix}.
\end{equation}
Here, too, $M$ and $(I-C+MC)D$ are irreducible.

@@

Here we assume a large haploid population of haploids that occupies $n$ demes, $A_1, \ldots, A_n$. In this case we directly use the notation of Eq.~\ref{eq:model}.
Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.
The population mean fitness is $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$.

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k M_{k,j}$, where $C_k \ge 0$, $M_{k,j} \ge 0$ and $\sum_{k=1}^{n}{M_{k,j}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $x'_k$ are given by Eq.~\ref{eq:model_sum}.
If the rows of $M$ are equal, that is, $M_{k,j}=m_k$ for all $j$,
then $m_k$ can be considered the relative population size of deme $A_k$;
this is the non-homogeneous extension to Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.
The transition matrix $M$ can account for other factors that determine the probability of non-homing juveniles from deme $A_j$ to migrate to deme $A_k$ -- for example, deme adjacency:
if demes are arranged in a circle, 
then $M$ can take the form
\begin{equation}
M = \begin{bmatrix}
p_0 & p_1 & \ldots & p_n \\
p_n & p_0 & \ldots & p_{n-1} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_n & \ldots & p_0
\end{bmatrix}.
\end{equation}
where $p_{-k} = p_{n-k}$, $p_k \ge 0$, and $\sum_{k=1}^{n}{p_k}=1$.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k);
\end{equation}
therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

>>>>

According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e. mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated. 
The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation~\citep{Liberman1986a}, migration~\citep{Feldman1986}, and dispersal rates~\citep{Altenberg1987}.
These separate principles were recently unified by~\citet{Altenberg2017}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism in which the mutation rate increases in individuals with poor fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population in which fitness is determined by the number of mutant alleles accumulated in the genotype.
The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
\bar{w} f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\label{eq:model_sum}
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (\ci - C + MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$ such that $D_k \ne D_j$ for $k \ne j$, and we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$\ci$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

The types $A_k$ can represent a single or multiple haploid genetic loci or non-genetic traits.
Importantly, type transmission is vertical and uni-parental (the type is transmitted from a single parent to the offspring) and is independent of the frequencies of the other types.
Specifically, this model precludes processes such as recombination, social learning, sexual outcrossing, and horizontal or oblique transmission, as these processes are frequency-dependent~\cite[pg.~54]{Cavalli-Sforza1981}.

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
Importantly, different types can have different rates.
That is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

In the following section we present three examples of the model (Eq.~\ref{eq:model}) for mutation and migration.

@@

Here we consider a large population of haploids and a trait determined by a single genetic locus with $n$ possible alleles $A_1, A_2, \ldots, A_n$ and corresponding fitness values $D_1, D_2, \ldots, D_n$.
The mutation rates $C_k$ of individuals with allele $A_k$ are potentially different;
specifically, with probability $1-C_k$, the allele $A_k$ does not mutate, and with probability $\frac{C_k}{n-1}$, the allele $A_k$ mutates to $A_j$ for any $j \ne k$.
This is an extension of a model studied by~\citet{Altenberg2017} that allows for the mutation rate of an individual, $C_k$, to depend on the identity of the allele $A_k$.

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation, $f'_k$ in the next generation is given by
\begin{equation}
f_k' = \big(1 - C_k\big) \frac{D_k}{\bar{w}} f_k + \frac{1}{n-1} \sum_{k \ne j}{C_j \frac{D_j}{\bar{w}} f_j},
\label{eq:mutation_model_1}
\end{equation}
for all $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k=1}^{n}{D_k f_k}$ is the population mean fitness.

This model is a special case of the general model (Eq.~\ref{eq:model}) where
\begin{equation} 
{M} = \frac{1}{n-1} \begin{bmatrix}
0 & 1 & 1 & \ldots & 1 \\
1 & 0 & 1 & \ldots & \vdots \\
1 & 1 & 0 & \ldots & \vdots \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \ldots & \ldots & \ldots & 0
\end{bmatrix},
\end{equation}
with zeros on the diagonal and $\frac{1}{n-1}$ elsewhere.
Note that here ${M}$ is irreducible.

@@

Again, we consider a large population of haploids, but here individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
Specifically, the fitness of individuals with $k$ mutant alleles is $D_k$,
And the probability $C_k$ that a mutation occurs in individuals with $k$ mutant alleles depends on $k$.
We assume a low mutation rate $C_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$,
or it is \emph{beneficial} with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{D_0}{\bar{w}} f_0 + \beta C_{1} \frac{D_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{D_n}{\bar{w}} f_n + \delta C_{n-1} \frac{D_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{D_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{D_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{D_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.

Therefore, setting 
\begin{equation}
{M} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots & 0\\
\delta & 1-\delta-\beta & \beta &  \ddots & 0\\
0 & \delta & 1-\delta-\beta & \ddots & 0 \\
\vdots & \ddots & \ddots & \ddots & \beta \\
0 & 0 & 0 & \delta & 1-\beta
\end{bmatrix},
\end{equation}
with the beneficial, neutral, and deleterious mutation probabilities on the three main digonals and zeros elsewhere, 
Eq.~\ref{eq:mutation_model_2} can be viewed as a special case of Eq.~\ref{eq:model}.
Here, too, ${M}$ is irreducible.

@@

In this case we consider a large population of haploids that occupy $n$ demes, $A_1, \ldots, A_n$.
Let the frequencies of individuals in deme $A_k$ be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k} = 1$.
The fitness of individuals in deme $A_k$ is $D_k$, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes -- this has been termed \emph{hard selection}~\citep{Wallace1975,Karlin1982}.

After reproduction, offspring of individuals from deme $A_k$ return to their parental deme with probability $1-C_k$, or migrate to a different deme $A_j$ with probability $C_k \cdot M_{j,k}$, where $C_k > 0$, $M_{j,k} \ge 0$, and $\sum_{k=1}^{n}{M_{j,k}} = 1$ for all $j=1, \ldots, n$.
Therefore, $1-C_k$ can be interpreted as a \emph{homing rate}.

Following selection and migration the new frequencies $f'_k$ are given exactly by Eq.~\ref{eq:model_sum}.
If columns of ${M}$ are identical
\begin{equation}
{M} = \begin{bmatrix}
m_1 & m_1 & \ldots & m_1 \\
m_2 & m_2 & \ldots & m_2 \\
\vdots & \vdots & \vdots & \vdots \\
m_n & m_n & \ldots & m_n
\end{bmatrix},
\end{equation}
with $\sum_{k=1}^{n}{m_k}=1$,
then $m_k$ can be considered the relative population size of deme $A_k$ --
this is the non-homogeneous extension of Deakin's \emph{homing model}~\citep{Deakin1966,Karlin1982}.

Similarly, if demes are arranged in a circle, for example around a lake, 
then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
and ${M}$ takes the form
\begin{equation}
{M} = \begin{bmatrix}
p_0 & p_1 & \ldots & p_{n-1} \\
p_{n-1} & p_0 & \ldots & p_{n-2} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_2 & \ldots & p_0
\end{bmatrix},
\end{equation}
where $p_k \ge 0$ and $\sum_{k=1}^{n}{p_k}=1$.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$,
and $u$ and $v$ be the corresponding left and right eigenvectors.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

\begin{corollary}[Reproductive value principle]
In the notation of Lemma~\ref{lemma:mfp},
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
where $[M]_k$ is the $k$-th column of $M$.

Therefore, increased transition from type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}
